Publishing Datasets about Building Designs \\Based on a Multilayer Building Ontology

### 2015-04-28 article-02-main.tex
Changed rules for simple datatypes

\section{Generating ifcOWL Ontologies From IFC Specifications}

@@

\begin{ontologyRule}[Namespaces, ifcOWL layers: all]
All constructs related purely to EXPRESS, STEP or IFC specifications must be belong to namespaces with prefixes \name{expr:}, \name{step:} and \name{ifc:} accordingly. At this moment the namespaces are defined as in Listing \ref{ifcOWL-namespaces}. Instead of \name{IFCXXX} there should be IFC versions, e.g. \name{IFC2x3} or \name{IFC4}.
\end{ontologyRule}

\begin{lstlisting}[caption={Namespace definitions},label=ifcOWL-namespaces]
@prefix expr: <http://drumbeat.cs.hut.fi/owl/EXPRESS#>
@prefix step: <http://drumbeat.cs.hut.fi/owl/STEP#>
@prefix ifc: <http://drumbeat.cs.hut.fi/owl/IFCXXX#>
\end{lstlisting}

\subsection{Components of an EXPRESS schema}
The IFC specification is written as an EXPRESS schema. An EXPRESS schema defines series of data types, functions and rules by using EXPRESS data definition language specified in ISO10303-11 \cite{wiki:express,noauthor:ifc-guide}.

There are six groups of data types: simple data types, entity data types, enumeration data types, defined (declared) data types, select data types and aggregation data types. All types must be declared obviously by using constructs TYPE and ENTITY except the built-in simple data types. Definitions and characteristics of data types are considered below. EXPRESS schemata also contain definitions of functions, rules and type constraints most of which cannot be reflected in OWL 2 at all.

\subsection{Simple Data Type Declarations}
\label{ifcOWL:simple-type}

Simple data types built in EXPRESS are: \name{STRING}, \name{BI\-NA\-RY}, \name{IN\-TE\-GER}, \name{REAL}, \name{NUM\-BER}, \name{BOO\-LEAN} and \name{LO\-GI\-CAL}. However, type \name{LOGICAL} which has three possible values \name{TRUE}, \name{FALSE} and \name{UNKNOWN} is considered by us as an enumeration data type (see \ref{ifcOWL:enum-types}). The same to type \name{BOOLEAN} which is like a primitive boolean data type but semantically close and met with the same frequency in IFC4 as \name{LOGICAL}.

On the other hand, there are two defined data types in IFC (not in EXPRESS) that also can be referred to the simple-data-type group: \name{Ifc\-Time\-Stamp} which is a so-called TimeSpan data type and can be easily converted to \name{xsd:date\-Time}, and \name{Ifc\-Globally\-Unique\-Id} which represents set of compressed Globally Unique Identifiers (GUIDs), i.e. strings with fixed length (see Listing \ref{ifcOWL-additional-simple-types}).

\begin{lstlisting}[caption={Two defined data types considered as simple data types},label=ifcOWL-additional-simple-types]
(* number of seconds elapsed since the beginning of the year 1970 *)
TYPE IfcTimeStamp = INTEGER; END_TYPE;

@@

\begin{ontologyRule}[Simple data types, ifcOWL layers: all]
All simple data types except \emph{BOOLEAN} and \emph{LOGICAL} are converted to OWL classes which have only one property \name{rdf:value}. The range of that property should be the most similar primitive XSD type supported by OWL 2 \cite{w3c:owl2-profiles} and recommended to use in Linked Data Platform Best Practices and Guideline \cite{w3c:ldp-best-practices}. For \name{STRING} and \name{Ifc\-Globally\-UniqueId} it is \name{xsd:\-string}, for \name{BINARY} -- \name{xsd:\-hexBinary}, for \name{INTEGER} -- \name{xsd:\-integer}, for \name{REAL} and \name{NUMBER} -- \name{xsd:\-double}, for \name{Ifc\-Time\-Stamp} -- \name{xsd:\-dateTime} (see example on Listing \ref{lst:ifcOWL-simple-types}).
\end{ontologyRule}

\begin{lstlisting}[caption={Simple data types as instances of \name{owl:Class}}, label=lst:ifcOWL-simple-types]
expr:STRING a owl:Class;
    rdfs:subClassOf [ a owl:Restriction;
                      owl:onProperty rdf:value;
                      owl:allValuesFrom xsd:double; ] .
ifc:IfcTimeStamp a owl:Class;
    rdfs:subClassOf [ a owl:Restriction;
                      owl:onProperty rdf:value;
                      owl:allValuesFrom xsd:dateTime; ]
\end{lstlisting}

The selected XSD types conform to the value spaces of the original EXPRESS and IFC data types better than other ones. For instance, type \name{xsd:double} is better than \name{xsd:decimal} as it is an IEEE 64-bit floating-point data type and supports scientific notation \cite{w3c:xsd,datapic:xsd}.

The reason of using class boxing technique is to make simple data types be instances of \name{owl:Class}. Instead of that they could be either simply replaced with primitive XSD data types, or defined as shown in the Listing , but in these cases they will be instances of \name{rdfs:Datatype}.  and cannot be used for union classes

\begin{lstlisting}[caption={Simple data types as instances of \name{rdfs:Datattype}},label=lst:ifcOWL-simple-types-2]
# if LEVEL_OF_DETAIL is higher than STANDARD
expr:STRING owl:equivalentClass [
    rdf:type        rdfs:Datatype ;
    owl:onDatatype  xsd:string ;
    owl:withRestrictions ()
]
\end{lstlisting}

Other considered conversion options were:
\begin{itemize}
\item To replace simple data types with the primitive XSD data types. But in this case all select data types (see \ref{subsec:ifcOWL:select-types}) cannot be defined by means of property \name{owl:unionOf} the range of which

all properties in any OWL profile except OWL Full must be either \name{owl:}
\end{itemize}

Some of simple data types may be specified together with optional size parameter, e.g. \name{STRING(lengthInBytes)} or \name{BINARY(sizeInBits)}. If \name{\$LEVEL\-\_OF\-\_DETAILS} is less than \name{STANDARD} (see \ref{subsec:multilayer-ontology-approach}), these parameters are ignored, i.e. \name{STRING(N)} is considered as equivalent to \name{STRING}. Otherwise additional constraints are added to the primitive data type by using property \name{owl:onDatatype} and data type facets \name{xsd:maxLength} or \name{xsd:length} \cite{w3c:owl2-syntax,w3c:xmlschema11-2} (see Listing \ref{ifcOWL:length-constraint}.

\begin{lstlisting}[caption={Length constraint},label=ifcOWL:length-constraint]
# if LEVEL_OF_DETAIL is higher than STANDARD
expr:STRING22 owl:equivalentClass [
    rdf:type        rdfs:Datatype ;
    owl:onDatatype  xsd:string ;
    owl:withRestrictions (
        [ xsd:length "22"^^xsd:integer ]
    )
]
\end{lstlisting}

\begin{lstlisting}[caption={},label=1]
expr:STRING rdf:type owl:Class;
    rdfs:subClassOf [
        owl:Restriction;
        owl:onProperty expr:hasValue;
        owl:allValuesFrom xsd:string;
    ].

ifc:ifcLabel rdf:type owl:Class;
    rdfs:subClassOf expr:STRING.
\end{lstlisting}

\begin{lstlisting}[caption={},label=1]

% http://answers.semanticweb.com/questions/229/how-to-define-a-own-notation-datatype

http://semantic-web-grundlagen.de/wiki/Guide_to_OWL_2_for_OWL_1_users#Datatypes

http://www2.informatik.hu-berlin.de/~wandelt/SW201213/9OWL2.pdf

Using that in OWL 2 DL is no problem. If you want to stay within one of the profiles though note that defining datatypes through such restrictions isn't allowed in any of them. So for increased interoperability (also with OWL 1 and RDFS) you could just use xsd:string instead. That should be perfectly fine with SKOS.

expr:STRING rdf:type rdfs:Datatype;
 owl:onDatatype xsd:string;
 owl:withRestrictions ().

ifc:ifcLabel rdf:type owl:Class;
    rdfs:subClassOf expr:STRING.
\end{lstlisting}

\subsection{Entity Data Type Declarations}

>>>>

\section{Generating ifcOWL Ontologies From IFC Specifications}
\label{sec:ifcOWL}

@@

\begin{ontologyRule}[Namespaces, ifcOWL layers: all]
All constructs related purely to EXPRESS, STEP or IFC specifications must be belong to namespaces with prefixes \name{expr:}, \name{step:} and \name{ifc:} accordingly. At this moment the namespaces are defined as in Listing \ref{lst:ifcOWL-namespaces}. Instead of \name{IFCXXX} there should be IFC versions, e.g. \name{IFC2x3} or \name{IFC4}.
\end{ontologyRule}

\begin{lstlisting}[caption={Namespace definitions},label=lst:ifcOWL-namespaces]
@prefix expr: <http://drumbeat.cs.hut.fi/owl/EXPRESS#>
@prefix step: <http://drumbeat.cs.hut.fi/owl/STEP#>
@prefix ifc: <http://drumbeat.cs.hut.fi/owl/IFCXXX#>
\end{lstlisting}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Components of an EXPRESS schema
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Components of an EXPRESS schema}
The IFC specification is written as an EXPRESS schema. An EXPRESS schema defines series of data\-types, functions and rules by using EXPRESS data definition language specified in ISO10303-11 \cite{wiki:express,noauthor:ifc-guide}.

There are six groups of data\-types: simple data\-types, entity data\-types, enumeration data\-types, defined (declared) data\-types, select data\-types and aggregation data\-types. All types must be declared obviously by using constructs TYPE and ENTITY except the built-in simple data\-types. Definitions and characteristics of data\-types are considered below. EXPRESS schemata also contain definitions of functions, rules and type constraints most of which cannot be reflected in OWL 2 at all.

\subsection{Simple Datatype Declarations}
\label{subsec:ifcOWL-simple-type}

Simple data\-types built in EXPRESS are: \name{STRING}, \name{BI\-NA\-RY}, \name{IN\-TE\-GER}, \name{REAL}, \name{NUM\-BER}, \name{BOO\-LEAN} and \name{LO\-GI\-CAL}. However, type \name{LOGICAL} which has three possible values \name{TRUE}, \name{FALSE} and \name{UNKNOWN} is considered by us as an enumeration data\-type (see \ref{ifcOWL:enum-types}). The same to type \name{BOOLEAN} which is like a primitive boolean data\-type but semantically close and met with the same frequency in IFC4 as \name{LOGICAL}.

On the other hand, there are two defined data\-types in IFC (not in EXPRESS) that also can be referred to the simple-data-type group: \name{Ifc\-Time\-Stamp} which is a so-called TimeSpan data\-type and can be easily converted to \name{xsd:date\-Time}, and \name{Ifc\-Globally\-Unique\-Id} which represents set of compressed Globally Unique Identifiers (GUIDs), i.e. strings with fixed length (see Listing \ref{lst:ifcOWL-additional-simple-types}).

\begin{lstlisting}[caption={Two defined data\-types considered as simple data\-types}, label=lst:ifcOWL-additional-simple-types]
(* number of seconds elapsed since the beginning of the year 1970 *)
TYPE IfcTimeStamp = INTEGER; END_TYPE;

@@

\begin{ontologyRule}[Simple data\-types, ifcOWL layers: all]
All simple data\-types except \emph{BOOLEAN} and \emph{LOGICAL} are the same as their most similar primitive XSD data\-types which are supported by OWL 2 \cite{w3c:owl2-profiles} and are among preferred standard data\-types in Linked Data Platform \cite{w3c:ldp-best-practices} (see Listing \ref{lst:ifcOWL-simple-types}). However, it is recommended to use directly XSD s like (e.g. \name{xsd:string}) instead of the EXPRESS data\-type constructs (e.g. \name{expr:STRING}).
\end{ontologyRule}

% For \name{STRING} and \name{Ifc\-Globally\-UniqueId} it is \name{xsd:\-string}, for \name{BINARY} -- \name{xsd:\-hexBinary}, for \name{INTEGER} -- \name{xsd:\-integer}, for \name{REAL} and \name{NUMBER} -- \name{xsd:\-double}, for \name{Ifc\-Time\-Stamp} -- \name{xsd:\-dateTime}.

\begin{lstlisting}[caption={Simple data types}, label=lst:ifcOWL-simple-types]
expr:STRING, ifc:IfcGlobalUniqueId owl:sameAs xsd:string .
expr:BINARY owl:sameAs xsd:hexBinary .
expr:INTEGER owl:sameAs xsd:integer .
expr:REAL, expr:NUMBER owl:sameAs xsd:double .
ifc:IfcTimeStamp owl:sameAs xsd:dateTime .
\end{lstlisting}

`Most similar XSD data\-types' mean the types from XSD schema that conform to the value spaces of the original EXPRESS and IFC data\-types better than other ones. For instance, type \name{xsd:double} is more similar to \name{REAL} than \name{xsd:decimal} as it is an IEEE 64-bit floating-point data\-type and supports scientific notation \cite{w3c:xsd,datapic:xsd}.

\begin{ontologyRule}[Simple data\-types with sizes, ifcOWL layers: all]If \name{STRING} and \name{BINARY} data\-types are specified with a size parameter, i.e. \name{STRING (max\-Length\-In\-Bytes)} or \name{BINARY(size\-In\-Bits)}, then they will be named in format \name{\textless type\_name \textgreater\textless size\textgreater} (e.g. \name{STRING22}). In layers \lite{} and \standard{} they are equivalent to data\-types without size. But in layer \advanced{}, in order to specify the size parameter, additional constraints must be added by using property \name{owl:onDatatype} and data\-type facets \name{xsd:maxLength} or \name{xsd:length} \cite{w3c:owl2-syntax,w3c:xmlschema11-2} (see example in Listing \ref{lst:ifcOWL-simple-types-with-size}).
\end{ontologyRule}

\begin{lstlisting}[caption={Size constraint of simple s},
label=lst:ifcOWL-simple-types-with-size]
# only for Lite, Standard layers
expr:STRING22 owl:sameAs expr:STRING .

# only for Advanced layer
expr:STRING22 owl:equivalentClass
    [   rdf:type        rdfs:Datatype ;
        owl:onDatatype  xsd:string ;
        owl:withRestrictions ( [ xsd:length "22"^^xsd:integer ] ) ] .
\end{lstlisting}

% \begin{lstlisting}[caption={},label=1]
% expr:STRING rdf:type owl:Class;
%     rdfs:subClassOf [
%         owl:Restriction;
%         owl:onProperty expr:hasValue;
%         owl:allValuesFrom xsd:string;
%     ].

% \begin{lstlisting}[caption={Simple data\-types as instances of \name{owl:Class}}, label=lst:ifcOWL-simple-types]
% expr:STRING a owl:Class;
%     rdfs:subClassOf [ a owl:Restriction;
%                       owl:onProperty rdf:value;
%                       owl:allValuesFrom xsd:double; ] .
% ifc:IfcTimeStamp a owl:Class;
%     rdfs:subClassOf [ a owl:Restriction;
%                       owl:onProperty rdf:value;
%                       owl:allValuesFrom xsd:dateTime; ]
% \end{lstlisting}

% The reason of using class boxing technique is to make simple data\-types be instances of \name{owl:Class}. Instead of that they could be either simply replaced with primitive XSD data\-types, or defined as shown in the Listing , but in these cases they will be instances of \name{rdfs:Datatype}.  and cannot be used for union classes

% \begin{lstlisting}[caption={Simple data\-types as instances of \name{rdfs:Datattype}},label=lst:ifcOWL-simple-types-2]
% # if LEVEL_OF_DETAIL is higher than STANDARD
% expr:STRING owl:equivalentClass [
%     rdf:type        rdfs:Datatype ;
%     owl:onDatatype  xsd:string ;
%     owl:withRestrictions ()
% ]
% \end{lstlisting}

% Other considered conversion options were:
% \begin{itemize}
% \item To replace simple data\-types with the primitive XSD data\-types. But in this case all select data\-types (see \ref{subsec:ifcOWL:select-types}) cannot be defined by means of property \name{owl:unionOf} the range of which

% all properties in any OWL profile except OWL Full must be either \name{owl:}
% \end{itemize}

% ifc:ifcLabel rdf:type owl:Class;
%     rdfs:subClassOf expr:STRING.
% \end{lstlisting}

% \begin{lstlisting}[caption={},label=1]

% % http://answers.semanticweb.com/questions/229/how-to-define-a-own-notation-data\-type

% http://semantic-web-grundlagen.de/wiki/Guide_to_OWL_2_for_OWL_1_users#Datatypes

% http://www2.informatik.hu-berlin.de/~wandelt/SW201213/9OWL2.pdf

% Using that in OWL 2 DL is no problem. If you want to stay within one of the profiles though note that defining data\-types through such restrictions isn't allowed in any of them. So for increased interoperability (also with OWL 1 and RDFS) you could just use xsd:string instead. That should be perfectly fine with SKOS.

% expr:STRING rdf:type rdfs:Datatype;
%  owl:onDatatype xsd:string;
%  owl:withRestrictions ().

% ifc:ifcLabel rdf:type owl:Class;
%     rdfs:subClassOf expr:STRING.
% \end{lstlisting}

\subsection{Defined Datatype Declarations}
\label{subsec:ifcOWL-defined-types}

% \item \textbf{Defined or declared data type}: This further specializes other data\-types — e.g., define a datatype \emph{PositiveInteger} that is of type integer with a positive value.

\subsection{Entity Datatype Declarations}
